\textit{Supplemental materials for}: How is experience transformed into memory?

### 2018-09-04 paper/supplemental_materials.tex
finished draft of supplemental materials

\subsection*{Feature importance analyses}
To determine the contribution of each feature to the structure of the video topic proportions, we conducted a ``leave one out'' anlaysis.  Specifically, we compared the original video topic trajectory (created using all hand-annotated features from the 1000 hand-annotated scenes spanning the \textit{Sherlock episode}; see \textit{Methods} for a full list of features) with video trajectories created using all but one type of feature.  We created temporal correlation matrices for each trajectory (using the topic proportions matrices) and correlated the upper triangles of each impoverished trajectory with the original feature-complete trajectory.  Observing a lower correlation between an impoverished trajectory (holding out a particular feature) and the feature-complete trajectory would suggest that the given feature played a more prominent role in shaping the structue of the feature-complete trajectory.  We found that hand-annotated narrative details provided the most structure to the feature-complete trajectory, whereas transcriptions of onscreen text provided the least structure (Fig.~\ref{fig:feature-importance}A).

@@

As described below, we next considered three more naunced measures of the memory performance and dynamics that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

@@

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (34) matrix. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participant that recalled an event as a function of serial position during encoding (Fig.~\ref{fig:list-learning}A).

@@

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of their serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then, for each recalled event (and each participant), we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero.  Finally, we averaged over the rows of the matrix to yield a 1 by 34 array representing the proportion of participants that recalled each event as a function of its serial position (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering refers to the extent to which participants group their recall responses according to encoding position~\citep{PolyEtal09}. For instance, if the participant recalled the video events in the exact order they occurred, this would yield a score of 1.  If the participant recalled the video events in reverse order, this would yield a score of 0.  Similarly, if the participant recalled the events in random order, this yield yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant (mean: 0.808, SEM: 0.022).  Overall, we found that participants with higher temporal clustering scores also tended to recall more events ($r = 0.62, p = 0.007$).

\paragraph{Semantic clustering scores.} Semantic clustering measures the extent to which participants grouped their recall responses according to semantic similarity~\citep{PolyEtal09}. Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant (mean: 0.813, SEM: 0.022).  We found that participants who exhibited stronger semantic clustering scores overall remembered more video events ($r = 0.55, p = 0.02$).

\subsection*{Additional measures of naturalistic memory}
To quantify the similarity between the video topic trajectory and individual recall topic trajectories, we considered several novel metrics.  First, we tested whether each participant's recall trajectory matched the video trajectory in a general sense. For each participant we filtered the video trajectory to only include the events that the participant remembered.  We then computed the root mean squared difference (RMSD) between the remaining video events and the (closest-matching) recalled events.  For example, if the topic vectors for a participant's recall event topic vectors matched the corresponding video event topic vectors exactly (and in order), the expected RMSD for those events would be 0.  However, if the participant's recall events did not perfectly match the video events, or if they were out of order, then the RMSD would be greater than 0.  To assess the significance of the match between the video and recall trajectories, we carried out a permutation procedure whereby, for each of 10000 repetitions, we circularly shifted the recall trajectories (in time) by a random amount and then re-computed the RMSD each time.  This yielded a distribution of ``null'' RMSD values for each participant.  The observed RMSD values reached significance (i.e., $p < 0.05$, reflecting that more than 95\% of the null RMSD values were greater than the observed RMSD value) for nine of the participants (3, 4, 8--13, and 17).  (For the remaining participants this test yielded $p$s$ < 0.25$.)  The observed RMSD values were also reliably correlated with hand-annotated memory performance across participants ($r = -0.57, p = 0.016$).  In other words, a closer match between the video and recall topic trajectories was related to better overall recall performance.

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event,  indexing how similar the recall event was to its matching video event.  We defined the precision as the average video-recall correlation across all of the events a participant recalled.  We found a strong correlation between hand-annotated memory performance and precision, suggesting that participants who remembered more events also remembered them more veridically ($r= 0.74, p = 0.0006$).

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events.
\textbf{JRM STOPPED HERE}
We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way). To this end, we computed a `distinctiveness' score for each participant (i.e., 1 - the correlation between a recall event and all non-matching video events).  Then, we averaged this measure over recall events within participant.  We found that participants with higher hand annotated memory performance also had higher distinctiveness scores (Pearson's $r(16) = 0.8, p = 0.0001$).

Lastly, we tested whether participants with better memory performance were also more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the actual order of events (filtering events that were actually recalled).  We found that participants who recalled more events also recalled more of them in order (Pearson's $r(16) = 0.5, p = 0.04$). In summary, we found that better memory performance was associated with more precise, distinctive and ordered recalls.

\subsection*{Additional measures of naturalistic memory}
\textbf{Precision.} This measure gives us an indication of the specific match between a video event and recall event, where values approaching 1 are highly precise and lower values are inprecise.  We defined ``precision'' as the correlation between a recall event and its matching (i.e., argmax) video event.

\textbf{Distinctiveness.} Distinctiveness quantifies how similar a recall event is to all non-matching video events.  It provides a metric of how uniquely a particular recall event describes a particular video event.  To compute it, a given recall event is correlated to all video events, the argmax is removed and the rest of the values are averaged. The resulting value is subtracted from 1 such that larger values indicate a more distinctive recall event.

\section*{Participant-level results referenced in the main text}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/corrmats}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{figure}

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/matchmats}
\caption{\small \textbf{Video-recall event model correlation matrices.} Each participant's video event by recall event correlation matrix.  The yellow boxes represent the maximum correlation in each column.}
\label{fig:matchmats}
\end{figure}

%\newpage
\renewcommand{\refname}{Supplemental references}
\bibliography{memlab}

>>>>

\subsection*{Feature importance analyses}
To determine the contribution of each feature to the structure of the video topic proportions, we conducted a ``leave one out'' analysis.  Specifically, we compared the original video topic trajectory (created using all hand-annotated features from the 1000 hand-annotated scenes spanning the \textit{Sherlock} episode; see \textit{Methods} for a full list of features) with video trajectories created using all but one type of feature.  We created temporal correlation matrices for each trajectory (using the topic proportions matrices) and correlated the upper triangles of each impoverished trajectory with the original feature-complete trajectory.  Observing a lower correlation between an impoverished trajectory (holding out a particular feature) and the feature-complete trajectory would suggest that the given feature played a more prominent role in shaping the structure of the feature-complete trajectory.  We found that hand-annotated narrative details provided the most structure to the feature-complete trajectory, whereas transcriptions of onscreen text provided the least structure (Fig.~\ref{fig:feature-importance}A).

@@

As described below, we next considered three more nuanced measures of the memory performance and dynamics that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

@@

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participant that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).

@@

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of their serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then, for each recalled event (and each participant), we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 34 array representing the proportion of participants that recalled each event as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering refers to the extent to which participants group their recall responses according to encoding position~\citep{PolyEtal09}. For instance, if a participant recalled the video events in the exact order they occurred, this would yield a score of 1.  If a participant recalled the video events in reverse order, this would yield a score of 0.  Similarly, if a participant recalled the events in random order, this yield yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant (mean: 0.808, SEM: 0.022).  Overall, we found that participants with higher temporal clustering scores also tended to recall more events ($r = 0.62, p = 0.007$).

\paragraph{Semantic clustering scores.} Semantic clustering measures the extent to which participants clustered their recall responses according to semantic similarity~\citep{PolyEtal09}. Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant (mean: 0.813, SEM: 0.022).  We found that participants who exhibited stronger semantic clustering scores overall remembered more video events ($r = 0.55, p = 0.02$).

\subsection*{Additional measures of naturalistic memory}
To quantify the similarity between the video topic trajectory and individual recall topic trajectories, we considered several novel metrics.  First, we tested whether each participant's recall trajectory matched the video trajectory in a general sense. For each participant we filtered the video trajectory to only include the events that the participant remembered.  We then computed the root mean squared difference (RMSD) between the remaining video events and the (closest-matching) recalled events.  For example, if the topic vectors for a participant's recall event topic vectors matched the corresponding video event topic vectors exactly (and in order), the expected RMSD for those events would be 0.  However, if the participant's recall events did not perfectly match the video events, or if they were out of order, then the RMSD would be greater than 0.  To assess the significance of the match between the video and recall trajectories, we carried out a permutation procedure whereby, for each of 10000 repetitions, we circularly shifted the recall trajectories (in time) by a random amount and then re-computed the RMSD each time.  This yielded a distribution of ``null'' RMSD values for each participant.  The observed RMSD values reached significance (i.e., $p < 0.05$, reflecting that more than 95\% of the null RMSD values were greater than the observed RMSD value) for nine of the participants (3, 4, 8--13, and 17).  (For the remaining participants this test yielded $0.05 \leq p < 0.25$.)  The observed RMSD values were also reliably correlated with hand-annotated memory performance across participants ($r = -0.57, p = 0.016$).  In other words, a closer match between the video and recall topic trajectories was related to better overall recall performance.

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We defined the precision as the average video-recall correlation across all of the events a participant recalled.  We found a strong correlation between hand-annotated memory performance and precision, suggesting that participants who remembered more events also remembered them more veridically ($r = 0.74, p = 0.0006$).

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recalled event as
\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant.  We found that participants with higher average distinctiveness scores tended to also have better hand-annotated memory performance ($r = 0.8, p = 0.0001$).

\paragraph*{Other order effects.}
We tested whether participants with better memory performance were also more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the order the events actually occurred in the video (considering in the analysis only the events that the participant recalled).  Participants who recalled more events also recalled more of them in order ($r = 0.5, p = 0.04$). In summary, we found that better memory performance was associated with more precise, distinctive, and ordered recalls.

\section*{Participant-level figures referenced in the main text}

@@

\begin{figure}[htp]
\centering
\includegraphics[width=\textwidth]{figs/corrmats}
\caption{\small \textbf{Recall trajectory temporal correlation matrices and event segmentation fits.} Each panel is in the same format as Figure~2E in the main text.  The yellow boxes indicate HMM-identified event boundaries.}
\label{fig:corrmats}
\end{figure}

\begin{figure}[htp]
\centering
\includegraphics[width=\textwidth]{figs/matchmats}
\caption{\small \textbf{Video-recall event correlation matrices.}  Each panel is in the same format as Figure~2G in the main text.  The yellow boxes mark the maximum correlation in each column.}
\label{fig:matchmats}
\end{figure}

\newpage
\renewcommand{\refname}{Supplemental references}
\bibliography{memlab}
